

% --------------------------------
% INTRODUCTION
% --------------------------------


\chapter{Introduction\label{chapter:intro}}
\begin{comment}

Guides:
    - Find more things into this guides list
    - Page limit 1-2 pages
    - Context, who needs, what is needed, why is it a problem
    - Research question and research methodologies
    - Summary of results
    - What roles each section will have
    - No subsections

TODO:
    [ ] Who is this thesis for
    [ ] Why should you read my thesis
    [ ] What is the research question and how it is answered
    [ ] How is this thesis organized, what is covered and what is deliberaly not covered and in what chapters (outline)

What to cover:
    - Motivate the user to read my thesis!
    - What is cybersecurity and why it's of paramount importance
    - What is SE?
    - Modern AI
        - Large Language Models
        - Emergence of ChatGPT and the like
    - Attacks
        - Deepfake synthetic content, videos, live voice morphing
        - Highly personalized phishing content (natural language processing)
        - Automated OSINT gathering
    - AI augments both attacks and countermeasures
    - Countermeasures
        - User awareness & training programs
        - Company policy & company culture
        - Real-time threat detection
        - Vulnerability detection
    - Emerging challenges
    - Evolution of SE tactics and their impact on organizations
    - Intro to AI and its applications in cybersecurity
    - Research ga in the literature regarding the intersection of SE and AI


Info from lecture materials:
    - Johdannon tarkoituksena on kertoa yleiskielisesti työn tavoite. Kerrotaan (kuten tiivistelmässäkin, mutta laveammin), mitä on tutkittu, miten on tutkittu ja mitä tuloksia on saatu. Jotta kysymyksenasettelu ja tulokset on lukijan helppo oikein tulkita on syytä aloittaa johdanto asettelemalla tutkimus asiayhteyteensä, esimerkiksi kertomalla aluksi, minkälaisessa yhteydessä tarkasteluun otettavat haasteet esiintyvät ja keiden on ratkaisuista tarkoitus hyötyä.

    - Johdannon pituus määräytyy suhteessa koko kirjoitelman pituuteen. Parisivuinen kirjoitus ei erikseen otsikoitua johdantoa kaipaa, sillä se itsessään on laajennettu tiivistelmä. Kymmensivuisen kirjoituksen johdanto voi olla vaikkapa sivun tai puolentoista mittainen. Pro gradu -tutkielman 50-70-sivuiseen kokonaisuuteen tuntuu 2-4-sivuinen johdanto kohtuulliselta.

    - Johdanto kertoo siis lyhyessä, yleistajuisessa muodossa koko kirjoitelman kysymyksenasettelun, juonen sekä tulokset ja johtopäätelmät. Tämän luettuaan lukija voi päätellä, haluaako syventyä asiaan tarkemmin lukemalla koko kirjoituksen.

\end{comment}

The widespread adoption of interconnected IT devices and services has transformed every aspect of human life, from personal communication to business operations, and this reliance seems to be ever-expanding. The digital revolution has created numerous opportunities but has also introduced significant vulnerabilities. Among these, social engineering poses a particularly grave threat to security and privacy.

Social engineering is the art and science of manipulating victims into divulging confidential information or performing actions that may or may not be in their best interests \citep{hadnagySocialEngineering2018}. Rather than looking for technical vulnerabilities, social engineering relies on human interaction and exploits weaknesses in human psychology \citep{wangDefiningSocialEngineering2020}.




Historically, social engineering relied heavily on human intuition and manual effort to deceive targets \citep{mitnickArtDeceptionControlling2003}. With the advent of modern artificial intelligence (AI), the landscape of social engineering is undergoing a  significant transformation, augmenting the sophistication and effectiveness of current and emerging attack methods \citep{fakhouriAIDrivenSolutionsForSocialEngineeringAttacks2024}.


This thesis explores the intersection of AI and social engineering and how contemporary AI technologies enhance the execution and impact of these attacks, and discusses the necessary actions to counter such advanced attacks. Several social engineering attack vectors and tools particularly relevant to the modern threat of AI were selected for in-depth analysis: spear phishing with the help of chatbots like ChatGPT, impersonation through deepfake-generated images, audio, and video, and the automated gathering and processing of intelligence data.



Based on a literature review and analysis of occurred social engineering incidents, the results indicate that contemporary countermeasures against social engineering attacks are ill-equipped to deal with the sophistication of AI-powered threats \citep{fakhouriAIDrivenSolutionsForSocialEngineeringAttacks2024, blauthArtificialIntelligenceCrime2022}. Thus, the urgent need for cybersecurity professionals to update their tools and strategies is evident; however, AI can also play a valuable role in this area.





%
% ORGANIZED
%

The rest of the thesis is structured as follows: Chapter \ref{chapter:definition} introduces social engineering, generative AI, and other essential concepts for further analysis. Chapter \ref{chapter:attacks} examines relevant attack vectors and tools, including spear phishing and deepfake impersonation. Chapter \ref{chapter:countermeasures} then discusses both technological and human-oriented countermeasures against these attacks. The effectiveness and viability of these measures are then assessed in Chapter \ref{chapter:evaluation}. Lastly, Chapter \ref{chapter:conclusions} summarizes key findings and implications for the future of social engineering defense.




%Despite advancements in cybersecurity measures, social engineering remains a persistent threat. High-profile data breaches and corporate espionage incidents underscore the efficacy of social engineering techniques. Often, sophisticated firewalls, encryption algorithms and intrusion prevention systems (IPS) and intrusion detection systems (IDS) can be rendered useless by a cleverly worded email or an unsuspecting phone call. While it's often said that the most secure computer is one that is not plugged in, with social engineering an attacker could persuade someone to plug it in and turn it on.

%This study seeks to investigate the diverse aspects of social engineering, focusing on its techniques, effects, and prevention measures. Grasping the concept of social engineering is essential, not just for cybersecurity experts but also for people and organizations in every sector and at all employment levels. The increasing sophistication of social engineering attakcks coupled with the integration of artificial intelligence (AI) and technologies such as deepfakes (deep learning, fake) pose a growing risk that transcends traditional security frameworks. Because of this, extra attention is paid to how AI is and might contribute to the field of social engineering, both in terms of attacks and defenses.

%Efficacy of social engineering training and awareness programs varies, with main challenges lying in user's interest and motivation towards the training and retention of learned data. Continued education, especially because social engineering is a constantly evolving field where the attackers never cease to improvise, is absolutely essential. Gamification, and users's inward motivation, go a long way in achievehing this.



%Currently, social engineering attacks are the biggest threat facing cybersecurity.


