

% --------------------------------
% INTRODUCTION
% --------------------------------


\chapter{Introduction\label{chapter:intro}}
\begin{comment}

Guides:
    - Find more things into this guides list
    - Page limit 1-2 pages
    - Context, who needs, what is needed, why is it a problem
    - Research question and research methodologies
    - Summary of results
    - What roles each section will have

TODO:
    [ ] Who is this thesis for
    [ ] Why should you read my thesis
    [ ] What is the research question and how it is answered
    [ ] How is this thesis organized, what is covered and what is deliberaly not covered and in what chapters (outline)

What to cover:
    - Motivate the user to read my thesis!
    - What is cybersecurity and why it's of paramount importance
    - What is SE?
    - Modern AI
        - Large Language Models
        - Emergence of ChatGPT and the like
    - Attacks
        - Deepfake synthetic content, videos, live voice morphing
        - Highly personalized phishing content (natural language processing)
        - Automated OSINT gathering
    - AI augments both attacks and countermeasures
    - Countermeasures
        - User awareness & training programs
        - Company policy & company culture
        - Real-time threat detection
        - Vulnerability detection
    - Emerging challenges
    - Motives for cybercrimes
        - Hard(er) to detect?
        - "Easy" wins?
    - Evolution of SE tactics and their impact on organizations
    - Intro to AI and its applications in cybersecurity
    - Research ga in the literature regarding the intersection of SE and AI


Info from lecture materials:
    - Johdannon tarkoituksena on kertoa yleiskielisesti työn tavoite. Kerrotaan (kuten tiivistelmässäkin, mutta laveammin), mitä on tutkittu, miten on tutkittu ja mitä tuloksia on saatu. Jotta kysymyksenasettelu ja tulokset on lukijan helppo oikein tulkita on syytä aloittaa johdanto asettelemalla tutkimus asiayhteyteensä, esimerkiksi kertomalla aluksi, minkälaisessa yhteydessä tarkasteluun otettavat haasteet esiintyvät ja keiden on ratkaisuista tarkoitus hyötyä.

    - Johdannon pituus määräytyy suhteessa koko kirjoitelman pituuteen. Parisivuinen kirjoitus ei erikseen otsikoitua johdantoa kaipaa, sillä se itsessään on laajennettu tiivistelmä. Kymmensivuisen kirjoituksen johdanto voi olla vaikkapa sivun tai puolentoista mittainen. Pro gradu -tutkielman 50-70-sivuiseen kokonaisuuteen tuntuu 2-4-sivuinen johdanto kohtuulliselta.

    - Johdanto kertoo siis lyhyessä, yleistajuisessa muodossa koko kirjoitelman kysymyksenasettelun, juonen sekä tulokset ja johtopäätelmät. Tämän luettuaan lukija voi päätellä, haluaako syventyä asiaan tarkemmin lukemalla koko kirjoituksen.

\end{comment}

The widespread and interconnected adoption of IT devices and services has transformed every aspect of human life, from personal communication to business operations, and this reliance seems to be ever-expanding. Although this digital revolution has opened up many opportunities, it has also brought about considerable vulnerabilities. One of the most sinister threats to security and privacy is social engineering.%, with FBI's Internet Crime Complaint Center (IC3) reporting year 2023's losses to organizations and individuals from business email compromise scams alone reaching a record \$2.9 billion\footnote{https://www.ic3.gov/Media/PDF/AnnualReport/2023\_IC3Report.pdf (accessed 2024-07-13)}.

Social engineering is the art and science of manipulating victims into divulging confidential information or performing actions that may or may not be in their best interests \citep{hadnagySocialEngineering2018}. Rather than looking for technical vulnerabilities, social engineering relies on human interaction and exploits weaknesses in human psychology \citep{wangDefiningSocialEngineering2020}.

With the advent of modern artificial intelligence (AI), the landscape of social engineering is undergoing significant transformation, augmenting the sophistication and effectiveness of social engineering attacks. Based on a literature review, this thesis explores the intersection of AI and social engineering and how contemporary AI technologies enhance the execution and impact of social engineering attacks, and discusses the necessary actions to counter such advanced attacks.

%For illustration, famous case studies are also examined.

Historically, social engineering relied heavily on human intuition and manual effort to deceive targets. Today, AI systems are more and more capable of automating and amplifying these deceptive practices, often with shocking alarming precision. From sophisticated spear phishing schemes utilizing technologies such as ChatGPT to deepfake-technologies creating highly convincing fake identities via audio, pictures and even video, AI-powered attacks represent a truly profound shift within social engineering, which we have only begun to witness.

%Certain social engineering attacks that are of particular interest when it comes to AI were chosen for more in-depth analysis, such as phishing, vishing (voice phishing), automated intelligence gathering, and deepfake-generated content, while leaving other attacks with less focus, such as dumpster diving, shoulder surfing, and tailgating.

Certain social engineering attacks and tactics that are of particular interest when it comes to the modern threat of AI to social engineering were chosen for more in-depth analysis, such as spear phishing and deepfake-generated content, while leaving other attacks with less focus, such as dumpster diving, shoulder surfing, and tailgating.

%The thesis discusses countermeasures, emphasizing the need to enhance current cybersecurity training and awareness programs to address the modern AI threat. The primary focus is on the human aspect of social engineering, while the technical details of AI tools and techniques are beyond the scope of this thesis.

%Countermeasures that are especially covered include how current cybersecurity training and awareness programs need to be augmented for the modern threat of AI. The major focus of this thesis is in the human element of social engineering, and the technical details of AI tools and techniques are outside of this thesis's scope.

%Next, key ideas and concepts in the field of social engineering are examined before moving on to the anatomy of a typical social engineering attack so that the actual attacks can be discussed in their proper context, all of this before we can finally delve into the countermeasures against the discsused attacks.

With the sophistication of AI-powered attacks, current countermeasures are becoming ever more ineffective, necessitating the urgent need for cybersecurity professionals to update their strategies and tools for cyber defense.

The rest of the thesis is organized as follows.
Chapter \ref{chapter:definition} provides an overview of social engineering, generative AI and other concepts necessary for further analysis.
Common social engineering attack vectors and tools, such as spear phishing and impersonation with deepfakes, and  how modern AI amplifies their effectiveness are examined in Chapter \ref{chapter:attacks}.
Next, Chapter \ref{chapter:countermeasures} discusses countermeasures against these attacks, both technology- and human-oriented.
The effectiveness and viability of the discussed countermeasures is evaluated in Chapter \ref{chapter:evaluation}.
Finally, Chapter \ref{chapter:conclusions} concludes the thesis, summarizing the key findings and implications for the future of social engineering defense.

%Chapter \ref{chapter:anatomy} delves into the various stages of a typical social engineering attack, highlighting the tactics and strategies employed by attackers. Chapter \ref{chapter:attacks} examines common attack methods and vectors, such as phishing, vishing (voice phishing), and pretexting, and explores how modern AI capabilities amplify their effectiveness.




%Despite advancements in cybersecurity measures, social engineering remains a persistent threat. High-profile data breaches and corporate espionage incidents underscore the efficacy of social engineering techniques. Often, sophisticated firewalls, encryption algorithms and intrusion prevention systems (IPS) and intrusion detection systems (IDS) can be rendered useless by a cleverly worded email or an unsuspecting phone call. While it's often said that the most secure computer is one that is not plugged in, with social engineering an attacker could persuade someone to plug it in and turn it on.

%This study seeks to investigate the diverse aspects of social engineering, focusing on its techniques, effects, and prevention measures. Grasping the concept of social engineering is essential, not just for cybersecurity experts but also for people and organizations in every sector and at all employment levels. The increasing sophistication of social engineering attakcks coupled with the integration of artificial intelligence (AI) and technologies such as deepfakes (deep learning, fake) pose a growing risk that transcends traditional security frameworks. Because of this, extra attention is paid to how AI is and might contribute to the field of social engineering, both in terms of attacks and defenses.

%By dissecting real-world case studies and theoretical underpinnings, this thesis will provide a comprehensive overview of social engineering, contributing to the broader effort of mitigating its risks.

%Efficacy of social engineering training and awareness programs varies, with main challenges lying in user's interest and motivation towards the training and retention of learned data. Continued education, especially because social engineering is a constantly evolving field where the attackers never cease to improvise, is absolutely essential. Gamification, and users's inward motivation, go a long way in achievehing this.



%Usually when IT experts talk about an information system, they focus on the technical parts, the devices. But most, if not all, IT systems are useless without at least some interface with the people using it. Thus, especially from the social engineering standpoint, it's more beneficial to think of an information system as comprising the technology and its users. With this in mind, it becomes evident that the attack surface a hacker has is vast, and like water flowing down a river, the hacker often chooses the path of least resistance. Time and time again it has been proven that the easiest, most cost-effective way in is through the human factor of the IT system. Bolstering just one facet of this system is like fixing a leak or a potential leak only in some places, and not looking at the whole situation holistically. Security and privacy are a 360 degree, 24/7 thing.

%Currently, social engineering attacks are the biggest threat facing cybersecurity.


%This thesis is organized as follows. First, social engineering is defined and some history explained. Next, the different stages of a social engineering attack are examined. Then multiple common attack methods are discussed along with how modern AI powers them up. After that, AI-powered countermeasures are examined. The final chapter conclues this thesis.



