

% --------------------------------
% ABSTRACT (English)
% --------------------------------

\begin{otherlanguage}{english}
\begin{abstract}

\begin{comment}

Guides:
    - Max about 100-200 words
        - Some other theses have almost fully covered the available area
    - If the abstract text is too full, is that demotivating to the reader?
    - Study material states that the abstract text is short, usually just one paragraph?
    - What has been studied
    - How it has been studied
    - What results have been observed

TODO:
    [x] What is the research question
    [x] What is social engineering, quick definition?
    [ ] How does modern AI augment SE attacks and countermeasures
    [ ] What is analyzed in this thesis
    [ ] Properly motivate the reader to read my thesis
    [ ] Explain that this is written in an easy-to-read manner?
    [ ] What results have been observed

What to cover:
    - What is SE?
    - Modern AI
    - Attacks
        - Deepfake synthetic content, videos, live voice morphing
        - Highly personalized phishing content (natural language processing)
        - Automated OSINT gathering
    - AI augments both attacks and countermeasures
    - Countermeasures
        - User oriented
            - User awareness & training programs
            - Company policy & company culture
        - Tech oriented
            - Real-time threat detection
            - Vulnerability detection
    - Emerging challenges

\end{comment}

Social engineering, a subdomain of cybersecurity, is the art and science of manipulating people into divulging confidential information or taking actions that may or may not be in their best interests. Traditionally, social engineering relied heavily on manual labor and human intuition, but with the advent of modern AI technologies, cybercriminals are able to craft highly targeted and effective social engineering campaigns, with novel unexpected twists, that are sometimes entirely automated.

%This thesis explores the evolving landscape of social engineering in the modern age, paying special attention to how technologies such as artificial intelligence (AI) are impacting it. While AI offers novel opportunities for enhancing cybersecurity measures, it simultaneously empowers malicious actors with sophisticated tools for crafting and executing highly targeted and effective social engineering campaigns that are sometimes almost fully automated.

%In contrast, the thesis also explores how contemporary countermeasures against these advanced social engineering attacks are no longer sufficient. By analyzing the versatile impact of AI on social engineering, this thesis hopes to provide a comprehensive understanding of the emerging challenges and opportunities, with the countermeasures examined being especially focused on the human element within social engineering.

This thesis explores the evolving landscape of AI on social engineering, focusing on attacks such as spear phishing, automated intelligence gathering and impersonation with deepfake generated content. It also covers countermeasures against these attacks, such as those aided by AI, and evaluates their effectiveness based on relevant literature and case studies. The results indicate that contemporary countermeasures against social engineering attacks are becoming increasingly ineffective, highlighting the urgent need for cybersecurity professionals to update their strategies for cyber defense.

%This thesis hopes to contribute to our understanding of AI's impact on social engineering and highlights the urgent need for cybersecurity professionals to adapt to these emerging threats.


%In-depth analysis of the various ways AI is being exploited to augment social engineering tactics, including the creation of highly believable synthetic media such as deepfake images, videos, and real-time voice morphing, the generation of personalized phishing messages through natural language processing, and the automated open source intelligence gathering processes are performed.


%Given the rapidly evolving nature of AI and its expanding capabilities, much of the content presented is speculative. These projections are grounded in an analysis of established social engineering attacks and observed trends in AI development.


%The findings reveal that AI-powered attacks are more convincing and successful than traditional attacks, but AI-driven defense mechanisms can improve detection rates, and AI can develop more engaging training material(?).



%While AI can help detect social engineering attacks, it does not mitigate the need for user training and awareness programs, quite the contrary, with AI-powered attacks the need for awareness and vigilance will likely grown even higher.


% Tiivistelmäteksti on lyhyt, yleensä yhden kappaleen mittainen (maksimissaan noin 100 sanaa) selvitys kirjoituksen tärkeimmästä sisällöstä: mitä on tutkittu, miten on tutkittu ja mitä tuloksia on saatu.




\end{abstract}
\end{otherlanguage}
